Lessons learned from bugs in models of human history

### 2020-08-05 paper.tex
more edits to figure captions and text

Even though the error has a limited effect on simulated data, the tutorial code has been
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test a non-demographic inference
method, and the model was used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
library, which removes the need to re-implement demographic models from scratch.
This is a ``standard library" of quality-controlled
models and simulation resources for a growing number of commonly studied species.
The resource is built around an open-source community development model
with rigorous code review and quality control procedures.
For example, when a contributor adds a new model it is not fully integrated
into the catalog until a second, entirely independent, implementation of the
model is provided by another developer. If the original and ``QC" versions
of the model are precisely equal, then we can have a reasonable degree
of confidence in correctness. It is through this QC process that we discovered
the misspecified model in \msprime's documentation.
If the required model is not present in \stdpopsim, then particular care should
be taken to validate the implementation.
This could include verification through code review or comparison with an
independent implementation of the model. If the model is of wider interest to
the community, we would encourage developers to contribute it to \stdpopsim:
at the very least, the quality-control procedures in place provide a strong
reassurance of correctness.

Secondly, regardless of whether the model has been implemented
locally or obtained from a resource like \stdpopsim,
basic statistical validation should \emph{always} be performed on the
results. Errors are all too easy to make, and an analysis of the
basic statistical properties of the simulations is essential
due diligence. For example, a brief analysis of $F_{ST}$ values
would likely would have prevented the error in~\citet{martin2017human}.
It is understandable that such analyses were not undertaken in this
case, as processing such a huge dataset (600,000 samples) was
highly challenging. Recent progress has made statistical analysis scale much
easier for large sample sizes~\citep{ralph2020efficiently}, but the problems would have been
apparent if smaller pilot simulations were undertaken and analysed.
Graphical inspection of demographic models can also help identify issues, especially if
visualization can be automated. The demography debugging tool in \msprime\
summarizes demographic events occurring over time, and various efforts are underway to facilitate
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.

@@

We computed the expected allele frequency spectrum using
\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
were computed from the expected AFS and verified using branch statistics from
the output of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
Demographic models were plotted using the \texttt{demography} package
(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
We used the original pipeline from~\citet{martin2017human} available from
\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
We updated the pipeline to run with the correct demographic parameters and more
recent versions of \msprime\ and \tskit, and the updated pipeline is
available at \url{https://github.com/apragsdale/PRS}.
Data and Python scripts to recreate Figures~1~and~2 can be found at
\url{https://github.com/jeromekelleher/msprime-model-errors}.
A full list of the GitHub repositories containing copies of the erroneous
model are also given here.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\citet{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian populations is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
    For each population in the violin plots, the original correlations from~\citet{martin2017human}
    are shown on the left, and correlations using the correct model are shown on the right.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}

>>>>

Even though the error has a limited effect on simulated data, the tutorial code has been
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test a non-demographic inference
method, such as inferring gene genealogies from sequencing data~\citep{kelleher2019inferring},
determining the age of a genetic variant~\citep{albers2020dating}, 
or studying the power of variant association tests~\citep{tong2020population},
and the model was used as an example of a complex population history.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
library, which removes the need to re-implement demographic models from scratch.
This is a ``standard library" of quality-controlled
models and simulation resources for a growing number of commonly studied species.
The resource is built around an open-source community development model
with rigorous code review and quality control procedures.
For example, when a contributor adds a new model it is not fully integrated
into the catalog until a second, entirely independent, implementation of the
model is provided by another developer. If the original and ``QC" versions
of the model are precisely equal, then we can have a reasonable degree
of confidence in correctness. It is through this QC process that we discovered
the misspecified model in \msprime's documentation.
However, \stdpopsim has QC-verified models available for only a handful of published
models and species, so it might not meet the needs of a particular simulation study.
If the required model is not present in \stdpopsim, then extra care should
be taken to validate the implementation.
This could include verification through code review or comparison with an
independent implementation of the model. If the model is of wider interest to
the community, we would encourage developers to contribute it to \stdpopsim:
at the very least, the quality-control procedures in place provide a strong
reassurance of correctness.

Secondly, regardless of whether the model has been implemented
locally or obtained from a resource like \stdpopsim,
basic statistical validation should \emph{always} be performed on the
results. Errors are all too easy to make, and an analysis of the
basic statistical properties of the simulations is essential
due diligence. For example, a brief analysis of $F_{ST}$ values
would likely would have prevented the error in~\citet{martin2017human}.
It is understandable that such analyses were not undertaken in this
case, as processing such a huge dataset (600,000 samples) was
highly challenging. Recent progress has made statistical analysis scale much
easier for large sample sizes~\citep{ralph2020efficiently}, so that $F_{ST}$,
heterozygosity, the allele frequency spectrum, or the distribution of coalescence
times can be rapidly computed. However, the problems would have been
apparent if pilot simulations with smaller sample sizes were undertaken and analysed,
as simulations and analysis of small samples sizes is very efficient.
Graphical inspection of demographic models can also help identify issues, especially if
visualization can be automated. The demography debugging tool in \msprime\
summarizes demographic events occurring over time, and various efforts are underway to facilitate
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.

@@

We computed the expected allele frequency spectrum using
\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
were computed from the expected AFS and verified using branch statistics from
the output of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
Demographic models were plotted using the \texttt{demography} package written for Python
(\url{https://github.com/apragsdale/demography}, ver.~0.0.3). The \texttt{demography} package
allows users to define a demographic model as a directed acyclic graph using
\texttt{networkx}~\citep{hagberg2008exploring} which can be plotted as in Figures 1A,B and 2A.
\texttt{demography} also can also translate the graph-based demography into \texttt{msprime} input
commands or simulate the allele frequency spectrum or two-locus statistics using 
\texttt{moments}~\citep{jouganous2017inferring, ragsdale2019models}
or \texttt{dadi}~\citep{gutenkunst2009inferring}.

For the analysis of polygenic risk, we used the original pipeline
from~\citet{martin2017human} available from
\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
We updated the pipeline to run with the correct demographic parameters and more
recent versions of \msprime\ and \tskit, and the updated pipeline is
available at \url{https://github.com/apragsdale/PRS}.
Data and Python scripts to recreate Figures~1~and~2 can be found at
\url{https://github.com/jeromekelleher/msprime-model-errors}.
A full list of the GitHub repositories containing copies of the erroneous
model are also given here.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\citet{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian populations is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Panels B and C show the distribution of true and inferred PRS after standardizing PRS across all individuals
    in the simulation (this is in contrast to Figure 5 in Martin et al., which plotted unstandardized PRS). Because
    the polygenic trait was simulated as a threshold trait, it is the relative distribution of PRS across indivuduals
    that determined cases and controls instead of absolute true or inferred PRS.
    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
    For each population in the violin plots, the original correlations from~\citet{martin2017human}
    are shown on the left, and correlations using the correct model are shown on the right.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}
